---
title: "Prediction on Fully Funded Loans from Lending Clubs"
subtitle: "MATH 5800 --- Course Project"
author: Wenjie Wang and Hao Li
date: "`r format(Sys.time(), '%d %B %Y')`"
header-includes:
    - \usepackage{bm}
    - \widowpenalties 1 150
bibliography: project_group01.bib
output:
    beamer_presentation:
        theme: "Singapore"
        toc: true
        fig_width: 7
        fig_height: 4
        fig_caption: true
        slide_level: 2
        keep_tex: false
---


```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
## function ipak: install and load multiple R packages.
## check to see if packages are installed.
## install them if they are not, then attach them to the search path.
ipak <- function (pkg) {
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg))
        install.packages(new.pkg, repos = "https://cloud.r-project.org")
    sapply(pkg, require, character.only = TRUE, quietly = TRUE)
    invisible(0L)
}

## install packages needed
needpack <- c("knitr", "kfigr", "glmnet", "pROC", "ggplot2")
ipak(needpack)

## get output format
outFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to")

## source code
source("R/functions.R")

## load the results
load("results/logis.RData")
load("results/rocQ3.RData")
load("results/rocQ1.RData")
load("results/rocQ2.RData")

## load the datasets
load("cleanData/rawDat3.RData")
load("cleanData/loanQ3.RData")

## initial processing
obj <- logis[[2L]]
colVec <- gg_color_hue(3)

## other temp results
rocDat1 <- as.data.frame(rocQ1[1:4])
rocDat2 <- as.data.frame(rocQ2[1:4])
rocDat3 <- as.data.frame(rocQ3[1:4])
aucQ1 <- as.numeric(rocQ1$auc)
aucQ2 <- as.numeric(rocQ2$auc)
aucQ3 <- as.numeric(rocQ3$auc)
ciQ3 <- paste(paste0(round(as.numeric(rocQ3$ci)[- 2L], 2), "%"),
              collapse = ", ")
fprQ1 <- with(rocDat1,
              100 - specificities[which(round(sensitivities, 2) == 90.00)[1]])
fprQ2 <- with(rocDat2,
              100 - specificities[which(round(sensitivities, 2) == 90.00)[1]])
fprQ3 <- with(rocDat3,
              100 - specificities[which(round(sensitivities, 2) == 90.00)[1]])

## temp function to set up index for figures and tables
noPrint <- function(lab) invisible(lab)

## test
fn = local({
  i = 0
  function(x) {
      i <<- i + 1
      if ("html" %in% outFormat) {
          return(paste('Figure ', i, ': ', x, sep = ''))
      }
      as.character(x)
  }
})

```

`r noPrint(figr("fig:lambda", TRUE, FALSE, type = "Figure"))`
`r noPrint(figr("fig:roc", TRUE, FALSE, type = "Figure"))`


# Introduction

## Background

- Once a loan is approved, it is not always fully funded.
- The loans being fully funded or not may depend on the characteristics of
  the loans themselves and the applicants.
- We can get over a hundred of characteristics for each loan from the lending
  club website.
- Taking the selected third quarter of 2016 data set as an example, there are
  around 100,000 approved loan records and the fully funded rate is 97.09%.

## Objective

- The features of the data motivate us:
    - use a regularized logistics regression model, which performs variable
      selection and estimation at the same time.

    - evaluate the prediction results using the receiver operating
      characteristic (ROC) curve and the area under curve (AUC) metrics.

- The objective of this project is to predict whether a loan would be fully
  funded using the training data set and evaluate the prediction results
  using the testing data sets.

# Data Processing

## Overview of Raw Data

- We selected the approved loan records of the third quarter of 2016 as the
  training data set and the first and second quarters of 2016 as the testing
  data sets.

- The raw data is available at the download section of the website
  of [Lending Club][lendingClub] in format of `.csv`.

- It contains totally `r ncol(rawDat3)` variables and `r nrow(rawDat3)`
  approved loan records.

## Generation of Response Variable

- Look up the variable dictionary and create a binary response indicating
  whether the loans were fully funded by comparing the amount committed
  by investors and the amount of the loans applied.

- The corresponding variables are named `funded_amnt_inv` and `loan_amnt`.

- The fully funded percent is `r paste0(round(mean(loanQ3$y) * 100, 2),"%")`,
  which is quite high compared with the not-fully funded percent. This means
  that our training data is highly imbalanced, and it is impossible to use
  ordinary models to fit the data or use accuracy to evaluate the prediction
  results.


## Transformation of Numeric Variables

- Most of the numeric variables, such as `loan_amnt`, `installment`, etc.,
  are non-negative due to their features.

- We applied 0-1 scaling transformation to them.


## Cleaning and Factorizing Text Variables

- Some text variables are not able to provide much information and should be
excluded from analysis.  For example, `desc` for loan description provided by
the borrower is almost all missing; `url` for the lending club page with
listing data is not accessible.

- Other character variables, such as `grade` for loan grade assigned by the
lending club does not have many levels. So they are taken as categorical
variables.



## Matching Job Title with Major Occupational Group

- Predictor `emp_title`: Job titles supplied by the borrower when applying
  for the loan.

- After the basic cleaning routine, it had totally
  `r length(unique(loanQ3$emp_title))` different levels.

- An appropriate and concise occupational group for those job titles was
  needed.

- The Occupational Classification System (OCS)
  provided by United States Department of Labor.

- Totally nine well-defined Major Occupational Group (MOG) from OCS.  For
  example, the [group A][mogA] is for professional and technical occupations.

---

- Term frequency and inverse document frequency (TF-IDF) is a popular
  text-mining technique used to quantify what a document is about.

- Term frequency (TF), how frequently a word occurs in a document measuring
  how important the word may be.

- Inverse document frequency (IDF) was designed to decrease the weight for
  commonly used words and increases the weight for words that are not used
  very much in a collection of documents. It can be combined with term
  frequency to calculate a term's TF-IDF.

---

- For each MOG, we considered every job title as a document containing words
  and computed TF-IDF score for each word in the original job title.

- For every loan record, we defined the final score of its job title to be
  the summation of the TF-IDF score of each word.

- Further, we assigned each job title to the MOG with the maximum score among
  the nine major occupational groups.

- If there was no unique maximum score, we assigned the job title to group,
  "others".


# Regularized Logistics Model

## Regularized Logistics Regression Model

- After data processing, over hundreds of predictors were included.

- The regularized model performs variable selection and estimation at
  the same time. In addition, regularization often prevents models from
  over-fitting and helps prediction.

- The regularized logistics regression model was applied and the LASSO type
  penalty proposed by @lasso was specified.

- The Estimates from the fitted model revealed very interesting results.


## Selection of Tuning Parameter

- The tuning parameter was selected by ten-folds cross-validation to be the
  largest value such that error was within 1 standard error of the minimum.

- The binomial deviance of the fitted model under different tuning parameter
  $\lambda$ was given in `r figr("fig:lambda", TRUE, FALSE, "Figure")`.

---

```{r fig:lambda, echo = FALSE, fig.width = 4.2, fig.height = 2.5, fig.cap = fn("Selection of tuning parameter."), fig.align = "center"}
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0)) # remove margin
plot(obj)
```

# Prediction

## Receiver Operating Characteristic (ROC) Curve

- The Receiver operating characteristic (ROC) curve: A graphical tool of
  illustrating the performance of a binary classifier system as its
  discrimination threshold varies.

- For the training set, loan records of Q3 in 2016:
    - With TPR being at least 90%, the smallest FPR was only around
      `r paste0(round(fprQ3, 2), "%")`.
    - The area under the curve (AUC) is `r paste0(round(aucQ3, 2), "%")`.

- For the test sets, loan records of Q1 and Q2 in 2016:
    - With TPR being at least 90%, the smallest FPR was about
      `r paste0(round(fprQ1, 2), "%")` and `r paste0(round(fprQ2, 2), "%")`,
      respectively.
    - The AUC was `r paste0(round(aucQ1, 2), "%")` and
      `r paste0(round(aucQ2, 2), "%")`, respectively.

---

```{r fig:roc, echo = FALSE, fig.width = 4.5, fig.height = 2, fig.cap = fn("The ROC curve of prediction from regularized logistics model for different quarters."), fig.align = "center"}
## Q1
p1 <- ggplot(rocDat1, aes(x = (100 - specificities) / 100,
                          y = sensitivities / 100)) +
    geom_line(color = colVec[1]) +
    geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "3313") +
    geom_hline(yintercept = 0.9, color = colVec[2], linetype = 2) +
    xlab("FPR") + ylab("TPR") + ggtitle("2016 Q1") + theme_bw() +
    theme(axis.text = element_text(size = 6),
          axis.title = element_text(size = 6),
          plot.title = element_text(size = rel(1)))
## Q2
p2 <- ggplot(rocDat2, aes(x = (100 - specificities) / 100,
                          y = sensitivities / 100)) +
    geom_line(color = colVec[1]) +
    geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "3313") +
    geom_hline(yintercept = 0.9, color = colVec[2], linetype = 2) +
    xlab("FPR") + ylab(NULL) + ggtitle("2016 Q2") + theme_bw() +
    theme(axis.text = element_text(size = 6),
          axis.title = element_text(size = 6),
          plot.title = element_text(size = rel(1)))
## Q3
p3 <- ggplot(rocDat3, aes(x = (100 - specificities) / 100,
                          y = sensitivities / 100)) +
    geom_line(color = colVec[1]) +
    geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "3313") +
    geom_hline(yintercept = 0.9, color = colVec[2], linetype = 2) +
    xlab("FPR") + ylab(NULL) + ggtitle("2016 Q3") + theme_bw() +
    theme(axis.text = element_text(size = 6),
          axis.title = element_text(size = 6),
          plot.title = element_text(size = rel(1)))
## combine output
multiplot(p1, p2, p3, cols = 3)

```

# Summary

## Summary

- We applied the regularized logistics regression model to the
  prediction problem of fully funded loans from lending clubs.

- The fitted model revealed very interesting results on the effect of
  predictors.

- The prediction results were evaluated using the ROC curve and the AUC, which
  suggested that the model did a fairly good job in prediction.

- We believe that a good model should provide us with not only reasonably
  good prediction but also inspiring interpretation of the variables of
  interest.

- The source document as well as all the **R** scripts are available at the
  GitHub repository named [math5800][github] we set up for this course.

# Acknowledgment

## Acknowledgment

The completion of the project would not be possible without the great effort
from each team member. It gave us much pleasure working together as a
group. We would also like to show our gratitude to Dr. Do who provided us
with the interesting lectures on financial programming and modeling.


# Reference

## Reference

[lendingClub]: https://www.lendingclub.com/info/download-data.action
[mogA]: http://www.bls.gov/ncs/ocs/ocsm/comMoga.htm
[glmnet]: https://CRAN.R-project.org/package=glmnet
[github]: https://github.com/wenjie2wang/math5800
[Rmarkdown]: http://rmarkdown.rstudio.com/
