---
title: "Backtesting and Comparing the Performance of Typical Portfolios"
subtitle: "MATH 5800 --- Course Project"
author: Wenjie Wang, Hao Li, and Catherine Payzant
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: |
    In this project, we reviewed the optimization algorithms for three
    popular portfolios: global minimum variance, mean-variance,
    and tangency covered in this course.
    We evaluated and compared the performance of these portfolios
    using the backtesting results from Quantopian. This project is
    motivated by a small Python algorithm for the minimum variance with
    constraints portfolio shared on the Quantopian community. We carry out a
    detailed discussion of the optimization algorithms and the backtesting
    results.
header-includes:
    - \usepackage{bm}
bibliography: project_group01.bib
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    number_sections: yes
    toc: no
  fontsize: 11pt
  classoption: letter
  documentclass: article
  html_document:
    theme: readable
    highlight: tango

---


```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
## function ipak: install and load multiple R packages.
## check to see if packages are installed.
## install them if they are not, then attach them to the search path.
ipak <- function (pkg) {
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg))
        install.packages(new.pkg, repos = "https://cloud.r-project.org")
    sapply(pkg, require, character.only = TRUE, quietly = TRUE)
    invisible(0L)
}

## install packages needed
needpack <- c("knitr", "kfigr", "quadprog")
ipak(needpack)

## get output format
outFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to")

## source code

## temp function to set up index for figures and tables
noPrint <- function(lab) invisible(lab)

## test
fn = local({
  i = 0
  function(x) {
      i <<- i + 1
      if ("html" %in% outFormat) {
          return(paste('Figure ', i, ': ', x, sep = ''))
      }
      as.character(x)
  }
})

```

`r noPrint(figr("efficient", TRUE, FALSE, type = "Figure"))`
`r noPrint(figr("fig:minVar", TRUE, FALSE, type = "Figure"))`
`r noPrint(figr("fig:meanVar", TRUE, FALSE, type = "Figure"))`
`r noPrint(figr("fig:tangency", TRUE, FALSE, type = "Figure"))`



# Introduction {#sec:intro}

In finance, an asset often means an investment instrument that can be traded.
Suppose one purchased an asset for $x_0$ dollars and sold it for $x_1$
dollars later, the relative difference in percentage, $(x_1 - x_0)/x_0$ is
called the rate of return on the asset.  Now consider an $N$ assets'
portfolio problem where $N$ is a finite positive integer. Let $r_i$,
$(i=1,2,\ldots,N)$, denote the rate of return on asset $i$ and assume it is
distributed as Gaussian distribution with mean $\mu_i$ and standard deviation
$\sigma_i$. Different assets are usually correlated with each other, which
implies another assumption that $\text{Cov}(r_i, r_j) = \sigma^2_{ij}$ for
all $i \ne j$. Let $\mathbf{r}=(r_1, r_2, \ldots, r_N)^T$ denote the expected
rate of returns and $\mathbf{\Sigma}=(\sigma^2_{ij})$ denote the
variance-covariance matrix representing the risk of all the assets considered
in the portfolio.  Suppose we assign weights $\mathbf{w} = (w_1, w_2, \ldots,
w_N)^T$ to those assets, then the expected return is $\mu =
\mathbf{r}^T\mathbf{w}$ and the corresponding risk is
$\sigma=\sqrt{\mathbf{w}^T\mathbf{\Sigma w}}$. The natural constraint on those
weights is that their summation has to be one. The weight taking negative
value means the short selling.


The objective of this project is to review some typical algorithms
determining the weights of any given set of assets so that the expected
return with risk level constrained is maximized.  How good their performance
will be in the real stock market is of great interest.  Back-testing those
algorithms in Quantopian platform provides an convenient way for evaluation
and comparison, where the formed portfolio can be automatically updated
periodically.


The rest of this report is organized as follows. In [Section 2](#sec:port),
we give a full description of the global minimum variance portfolio
in [Section 2.1](#sec:minVar), mean-variance portfolio
in [Section 2.2](#sec:meanVar), tangency portfolio
in [Section 2.3](#sec:tangency), and efficient frontier
in [Section 2.4](#sec:frontier).  In [Section 3](#sec:backtest), we present
the backtesting results of the implemented algorithms with some discussion
and plots.  We conclude the report with some summary and more discussion
in [Section 4](#sec:summary).


# Portfolio Forms {#sec:port}

## Global Minimum Variance Portfolio {#sec:minVar}

The global minimum variance (GMV) portfolio aims to minimize the overall risk
globally, which can be rewritten into a quadratic programming problem:
$$
\min_{\mathbf{w}} \mathbf{w}^T\mathbf{\Sigma}\mathbf{w},
~\text{s.t.}~\mathbf{w}^T\mathbf{1} = 1,
$$
where the only constraint is that the summation of the weights of assets in the
portfolio is one.


The GMV portfolio is designed to minimize overall risk. However, in reality,
investors usually tend to take more risk for possibly more returns.


## Mean-Variance Portfolio {#sec:meanVar}

The basic GMV portfolio can be easily extended by imposing extra constraints
on the objective function.  In this subsection, we briefly review the
Markowitz mean-variance portfolio.


Recall that $r_i$ is the expected rate of return on asset $i$.
Compared with GMV, the extra constraint that mean-variance portfolio
has is: $$\mathbf{w}^T \mathbf{r} - K_0 \ge 0,$$
where $\mathbf{r} = (r_1, r_2, \ldots, s_N)$ and $K_0$ is a
specified constant.


## Tangency Portfolio {#sec:tangency}

The tangency portfolio aims to maximize the ratio of excess return
($\mathbf{r}^T\mathbf{w} - r_f$) to portfolio volatility and represents a
trade-off between return and risk. The maximization problem is written as:
$$
\arg \max_{\mathbf{w}} \frac{\mathbf{r}^T\mathbf{w} - r_f}
{\sqrt{\mathbf{w}^T\mathbf{\Sigma}\mathbf{w}}},~ \text{s.t.}~
\mathbf{w}^T\mathbf{1} = 1,
$$
where $r_f$ is risk-free rate.

We may show that the maximization problem can be rewritten into a quadratic
programming problem.  Let $\mathbf{w}_0$ denotes the weight vector of the
tangency portfolio. Note that $\mathbf{r}^T \mathbf{w}_0 - r_f = \mathbf{r}^T
\mathbf{w}_0 - r_f \mathbf{1}^T\mathbf{w}_0$ since
$\mathbf{1}^T\mathbf{w}_0 = \mathbf{w}_0^T\mathbf{1} = 1$.  Then the maximum of the
target function can be rewritten as
$$
\frac{(\mathbf{r} - r_f \mathbf{1})^T \mathbf{w}_0}
{\sqrt{\mathbf{w}_0^T\mathbf{\Sigma}\mathbf{w}_0}}
= \frac{(\mathbf{r} - r_f \mathbf{1})^T \mathbf{w}_k}
{\sqrt{\mathbf{w}_k\mathbf{\Sigma}\mathbf{w}_k}},
$$
where $\mathbf{w}_k = k \mathbf{w}_0$, for some $k > 0$.
The maximum of the target function does not depend on the choice of $k$, which means
that we may find any $\mathbf{w}_k$ that maximizes the target function and
further re-scale them to satisfy the constraint:
$\mathbf{w}^T\mathbf{1} = 1$.
Therefore, the maximization problem of finding $\mathbf{w}_k$ is equivalent
to the following quadratic problem:
$$
\min_{\mathbf{w}} \mathbf{w}^T\mathbf{\Sigma}\mathbf{w},~
\text{s.t.}~ (\mathbf{r} - r_f \mathbf{1})^T \mathbf{w} = c_0,
$$
where $c_0$ is an arbitrary positive constant.  Therefore, the original
maximization problem is equivalent to the quadratic programming problem of
finding one weight vector $\mathbf{w}_k$, and scaling
$\mathbf{w}_k$ to $\mathbf{w}_0$.


## Efficient Frontier {#sec:frontier}

Following the framework originally developed by @markowitz1952, portfolios
maximizing expected return under certain given level of risk or equivalently
minimizing risk subject to a target expected return is defined as efficient
portfolios. The efficient frontier consists of all efficient portfolios.


Thus, the naive constructing method of efficient frontier is by obtaining the mean
variance portfolios subject to different target returns.  Alternatively, it
can be shown that any minimum variance portfolio can be represented as a
convex combination of any two minimum variance portfolios with different
target expected returns. If the expected return of the resulting portfolio is
greater than the expected return of the global minimum variance portfolio, then the
portfolio is an efficient frontier portfolio. (Otherwise, it is not.)
Therefore, we can also get the efficient frontier based on the global minimum
variance portfolio and tangency portfolio, both of which are known to be
efficient.


We constructed one sample efficient frontier shown in
`r figr("efficient", TRUE, FALSE, "Figure")`
from the simulated data for 20 assets.
Each asset had 60 simulated rate of returns from normal distribution
with mean 0.005 and standard deviation 0.015.
For simplicity, no specific correlation was imposed on these assets.
In addition, the expected returns and risk of those three portfolios were
plotted in `r figr("efficient", TRUE, FALSE, "Figure")`.
Clearly, global minimum variance, mean-variance, and tangency portfolio all
lie on the efficient frontier.


```{r efficient, echo = FALSE, fig.width = 6.5, fig.height = 3.5, fig.cap = fn("Sample efficient frontier."), fig.align = "center"}
## Tangency Portfolio ----------------------------------------------
## sample setting
na <- 20L                               # number of assets
ns <- 60L                               # number of observations
set.seed(1216)                          # important for reproducibility
retn <- matrix(rnorm(ns * na, mean = 0.005, sd = 0.015), nrow = ns)
mu <- colMeans(retn)                    # expected returns
rf <- 0.0001                            # riskfree rate (about 2.5% pa)
mu2 <- mu - rf                          # excess means
Q <- cov(retn)                          # covariance matrix

## obtain the optimized weights by quadratic programming
res_tan <- solve.QP(Dmat = Q, dvec = rep(0, na), Amat = matrix(mu2, ncol = 1),
                    bvec = 1, meq  = 1L)
w_tan <- prop.table(res_tan$solution)
retn_tan <- crossprod(w_tan, mu)
se_tan <- as.vector(sqrt(crossprod(w_tan, Q %*% w_tan))) # standard error
sr <- crossprod(w_tan, mu2) / se_tan  # compute sharpe ratio

## minimal variance portfolio
res_min <- solve.QP(Dmat = 2 * Q, dvec = rep(0, na),
                    Amat = matrix(1, nrow = na, ncol = 1),
                    bvec = 1, meq = 1L)
w_min <- res_min$solution
retn_min <- crossprod(w_min, mu)
se_min <- sqrt(as.vector(crossprod(w_min, Q %*% w_min)))

## mean-variance
eps <- 0.007
res_meanVar <- solve.QP(Dmat = 2 * Q, dvec = rep(0, na),
                        Amat = cbind(matrix(1, nrow = na, ncol = 1), mu),
                        bvec = c(1, eps), meq = 1L)
w_meanVar <- res_meanVar$solution
retn_meanVar <- crossprod(w_meanVar, mu)
se_meanVar <- sqrt(as.vector(crossprod(w_meanVar, Q %*% w_meanVar)))

## construct efficient frontier from minimal variance and tangency portfolio
retnVec <- c(retn_min, retn_tan)
seVec <- c(se_min, se_tan)
varMat <- diag(seVec ^ 2)
varMat[1, 2] <- varMat[2, 1] <- as.vector(crossprod(w_min, Q %*% w_tan))
alpha <- seq(from = 1, to = - 1, by = - 0.1)
nAlpha <- length(alpha)
tmpMat <- cbind(alpha, 1 - alpha)
retn_eff <- tmpMat %*% retnVec
var_eff <- sapply(alpha, function(a) {
    crossprod(c(a, 1 - a), varMat %*% c(a, 1 - a))
})

## hue function
gg_color_hue <- function(n) {
    hues <- seq(15, 375, length = n + 1)
    grDevices::hcl(h = hues, l = 65, c = 100)[seq_len(n)]
}
colVec <- gg_color_hue(3)

## plot the tangency and minimal variace portfolio, on efficient portfolio
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0)) # remove margin
plot(sqrt(var_eff), retn_eff, type = "l",
     xlab = expression(sigma[p]), ylab = expression(r[p]))
points(se_min, retn_min, col = colVec[1], pch = 20, cex = 1.5)
points(se_meanVar, retn_meanVar, col = colVec[2], pch = 20, cex = 1.5)
points(se_tan, retn_tan, col = colVec[3], pch = 20, cex = 1.5)
abline(b = (retn_tan - rf) / se_tan, a = rf, lty = 2, col = "gray", lwd = 1.5)
legend("bottomright", col = c("black", "gray", colVec),
       pch = c(NA, NA, rep(20, 3)), lty = c(1, 2, rep(NA, 3)),
       lwd = c(1, 1.5, rep(NA, 3)), cex = 0.85,
       legend = c("Efficient Frontier", "Captial Market Line (CML)",
                  "Global Minimum Variance", "Mean-Variance", "Tangency"))

```


# Backtesting {#sec:backtest}

We backtested the portfolios discussed in [Section 2](#sec:port) and compared
their performance in [Quantopian][quantopian]. The stocks of four famous
companies, *AAPL* (Apple Inc.), *AMZN* (Amazon Inc.), *GOOG* (Alphabet Inc.),
and *FB* (Facebook Inc.) were selected as sample stocks for backtesting.  All
the portfolios was implemented in **Python** with the help of Quantopian help
document and community.  The main body of function `initialize(context)` is
given in the following code chunk.

```{r initial, engine = "python", eval = FALSE}
def initialize(context):
    context.stocks = [sid(24),     # Apple Inc.
                      sid(16841),  # Amazon.com Inc.
                      sid(46631),  # Alphabet Inc. (Google's parent company)
                      sid(42950)]  # facebook Inc
    ## allocate weights of assets one hour after the market opens
    schedule_function(allocate, date_rules.every_day(),
                      time_rules.market_open(minutes = 60))
    ## trade one hour after the market opens every week
    schedule_function(trade, date_rules.week_start(days_offset = 1),
                      time_rules.market_open(minutes = 60))
    ## record leverage every day
    schedule_function(record_leverage, date_rules.every_day())
```


The selected stocks were save in `context.stocks`.  The function `allocate`
determined the weight of each stock at 10:30 AM on every market day.  The
function `trade` made weekly trades at 10:30 AM on the first day when the
market opens in that week.  For each asset, the trading weight was its
average weight from last five daily allocations.


For a fair comparison, the trading time and rules were kept the same for all
of the portfolios. The initial capital was set to be 10,000 USD. The time
period for the back-test was from 2014/06/01 to 2016/11/01.  Price history of
the last week in minutes was considered in function `data.history`.  The stock
market opens at 9:30 AM and closes at 4:00 PM. Thus one weekday produces 6.5
$\times$ 60 = 390 history prices for every minute. Then the sample rate of
returns were calculated by the relative difference between current prices and
last prices in percentage. The expected returns were estimated by the mean of
sample returns. The corresponding code is given as follows:


```{r getHistory, engine = "python", eval = FALSE}
prices = data.history(context.stocks, 'price', 5 * 390, '1m')
ret = prices.pct_change()[1 : ].as_matrix(context.stocks)
ret_mean = prices.pct_change().mean()
```


The only difference was the determination of weights of assets in the
portfolio as discussed in last section.  The main part of function `trade` is
given as follows:


```{r trade, engine = "python", eval = FALSE}
def trade(context, data):
    ## only trade when there is no open order
    if get_open_orders():
        return
    ## order to target weights
    for i, stock in enumerate(context.stocks):
        order_target_percent(stock, allocation[i])
    ## record weights for each stock for output
    record(AAPL = allocation[0])
    record(AMZN = allocation[1])
    record(GOOG = allocation[2])
    record(FB   = allocation[3])
```


We suppress all the remaining code to make this report concise and easy to
follow.  The complete **Python** scripts for these portfolios are available
at the [GitHub repository][github] we set up for this course.


`r figr("fig:minVar", TRUE, FALSE, "Figure")`,
`r figr("fig:meanVar", TRUE, FALSE, "Figure")`,
and `r figr("fig:tangency", TRUE, FALSE, "Figure")`
represents the results of the global minimum variance,
mean variance and the tangency portfolio, respectively.


```{r fig:minVar, echo = FALSE, results = "asis", fig.align = "center"}
src_tag <- "figs/globalMinVar.png"
caption <- "Backtesting results of the global minimum variance portfolio."
lab_num <- figr("fig:minVar", TRUE, FALSE, type = "Figure")
caption_tag <- paste0(lab_num, ": ", caption)

if ("html" %in% outFormat) {
    cat(paste("<div class='figure' style='text-align: center'>",
              "<img src=", src_tag,
              "alt=", caption_tag, "/>",
              "<p class='caption'>", caption_tag, "</p></div>"))
} else {
    cat("![", caption, "][fig:globalMinVar]", sep = "")
}
```

From `r figr("fig:minVar", TRUE, FALSE, "Figure")`, the GMV portfolio yielded
94.9% total returns at the end, which was much better than the benchmark by
gaining 79.7% more. The total return curve seem to be always above the
benchmark.  Also, the volatility and maximum draw-down rate were the smallest
among the three portfolios we considered here.

```{r fig:meanVar, echo = FALSE, results = "asis", fig.align = "center"}
src_tag <- "figs/meanVariance.png"
caption <- "Backtesting results of the mean-variance portfolio."
lab_num <- figr("fig:meanVar", TRUE, FALSE, type = "Figure")
caption_tag <- paste0(lab_num, ": ", caption)

if ("html" %in% outFormat) {
    cat(paste("<div class='figure' style='text-align: center'>",
              "<img src=", src_tag,
              "alt=", caption_tag, "/>",
              "<p class='caption'>", caption_tag, "</p></div>"))
} else {
    cat("![", caption, "][fig:meanVar]", sep = "")
}
```

For the mean-variance portfolio, the total returns at the end of the
back-testing was 70.3% of the initial investment. While the returns of the
benchmark (SPY) was just 15.2%. Therefore, the mean-variance portfolio also
did a better job than the benchmark algorithm by gaining another 55.1% of
total returns at the end.

```{r fig:tangency, echo = FALSE, results = "asis"}
src_tag <- "figs/tangency.png"
caption <- "Backtesting results of tangency portfolio."
lab_num <- figr("fig:tangency", TRUE, FALSE, type = "Figure")
caption_tag <- paste0(lab_num, ": ", caption)

if ("html" %in% outFormat) {
    cat(paste("<div class='figure' style='text-align: center'>",
              "<img src=", src_tag,
              "alt=", caption_tag, "/>",
              "<p class='caption'>", caption_tag, "</p></div>"))
} else {
    cat("![", caption, "][fig:tangency]", sep = "")
}
```

From `r figr("fig:tangency", TRUE, FALSE, "Figure")`, the total returns
produced by the tangency portfolio at the end of the back-testing was 106.6%
of the initial investment. Hence, it gave the largest returns among this
three portfolios.  The total returns gained at the end was about seven times
of the returns from the benchmark algorithm. Although it did not always
produce a better total returns than the benchmark at every time point, its
maximum draw-down was close to the GMV portfolio. However, it has the largest
volatility.


# Summary and Discussion {#sec:summary}

In summary, we reviewed global minimum variance portfolios, mean-variance
portfolios, and tangency portfolios that we learned from this course, and
further compared their performance by back-testing in Quantopian. The
backtesting results suggested that the tangency portfolio gave the largest
total returns at the end, while the global minimum variance portfolio
produced a consistently better total returns than the benchmark portfolio
over time. Also, its maximum drawdown rate was the smallest among those
portfolios.  All of these portfolios outperformed the benchmark algorithm
from Quantopian, though the performance of one certain portfolio is subject
to the selection of stocks, follow-up time period, etc.

The report is mainly powered by **[Rmarkdown][Rmarkdown]**. The source
document as well as the **Python** scripts used for backtesting in Quantopian
platform are available at the GitHub repository named [math5800][github] we
set up for this course.


# Acknowledgment {-}

The completion of the project would not be possible without the great effort
that our lovely team members have taken. It gave us much pleasure working
together as a group. We would also like to show our gratitude to Dr. Do who
introduced us the interesting lectures on financial programming and modeling,
and whose broad programming knowledge will have lasting effect.



# Reference {-}


[quantopian]: https://www.quantopian.com
[fig:meanVar]: figs/meanVariance.png
[fig:globalMinVar]: figs/globalMinVar.png
[fig:tangency]: figs/tangency.png
[github]: https://github.com/wenjie2wang/math5800
[Rmarkdown]: http://rmarkdown.rstudio.com/
