---
title: "Backtesting and Comparing the Performance of Typical Portfolios"
subtitle: "MATH 5800 --- Course Project"
author: Wenjie Wang, Hao Li, and Catherine Payzant
date: "`r format(Sys.time(), '%d %B %Y')`"
header-includes:
    - \usepackage{bm}
    - \widowpenalties 1 150
bibliography: project_group01.bib
output:
    beamer_presentation:
        theme: "Singapore"
        toc: true
        fig_width: 7
        fig_height: 4
        fig_caption: true
        slide_level: 2
        keep_tex: false
---


```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
## function ipak: install and load multiple R packages.
## check to see if packages are installed.
## install them if they are not, then attach them to the search path.
ipak <- function (pkg) {
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg))
        install.packages(new.pkg, repos = "https://cloud.r-project.org")
    sapply(pkg, require, character.only = TRUE, quietly = TRUE)
    invisible(0L)
}

## install packages needed
needpack <- c("knitr", "kfigr", "quadprog")
ipak(needpack)

## get output format
outFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to")

## source code

## temp function to set up index for figures and tables
noPrint <- function(lab) invisible(lab)

## test
fn = local({
  i = 0
  function(x) {
      i <<- i + 1
      if ("html" %in% outFormat) {
          return(paste('Figure ', i, ': ', x, sep = ''))
      }
      as.character(x)
  }
})

```

`r noPrint(figr("efficient", TRUE, FALSE, type = "Figure"))`
`r noPrint(figr("fig:MVWC", TRUE, FALSE, type = "Figure"))`
`r noPrint(figr("fig:minVar", TRUE, FALSE, type = "Figure"))`
`r noPrint(figr("fig:tangency", TRUE, FALSE, type = "Figure"))`


# Introduction

## Modern portfolio theory

- Objective: Use algorithms to determine weights of assets so the
  expected return is maximized given a certain level of risk.
- Quantopian was used to evaluate the results.


## Notations

- Consider an $N$ assets' portfolio problem where $N$ is a finite
  positive integer.
- Rate of return on asset $i$: $r_i$, $(i=1,2,\ldots,N)$.
- Mean $\mu_i$ and standard deviation $\sigma_i$.
- $\text{Cov}(r_i, r_j) = \sigma^2_{ij}$ for all $i \ne j$.
- Weights $\mathbf{w} = (w_1, w_2, \ldots, w_N)^T$.
- Expected return is then $\mu = \mathbf{r}^T\mathbf{w}$ and
  corresponding risk is
  $\sigma=\sqrt{\mathbf{w}^T\bm{\Sigma}\mathbf{w}}$, where
  $\mathbf{r}=(r_1, r_2, \ldots, r_N)^T$, and
  $\bm{\Sigma}=(\sigma^2_{ij})$.


# Typical Portfolios

## Global Minimum Variance Portfolio


- The global minimum variance (GMV) portfolio aims to minimize the overall risk
globally.
- A quadratic optimization problem:
$$
\min_{\mathbf{w}} \mathbf{w}^T\bm{\Sigma}\mathbf{w},
~\text{s.t.}~\mathbf{w}^T\mathbf{1} = 1,
$$
where the only constraint is that the summation of the weights of assets in the
portfolio is one.

- The GMV portfolio is designed to minimize overall risk.

- In reality, investors usually tend to take more risk for possibly more returns.

## Mean-Variance Portfolio

- The basic GMV portfolio can be easily extended by imposing extra constraints
on the objective function.

- Compared with GMV, the extra constraint that mean-variance portfolio
has is: $$\mathbf{w}^T \mathbf{r} - K_0 \ge 0,$$
where $\mathbf{r} = (r_1, r_2, \ldots, s_N)$ and $K_0$ is a
specified constant.

- A quadratic optimization problem:
$$
\min_{\mathbf{w}} \mathbf{w}^T\bm{\Sigma}\mathbf{w},
~\text{s.t.}~\mathbf{w}^T\mathbf{1} = 1, \mathbf{w}^T \mathbf{r} \ge K_0
$$

## Tangency Portfolio

- The tangency portfolio aims to maximize the ratio of excess return
($\mathbf{r}^T\mathbf{w} - r_f$) to portfolio violatility and represents a
trade-off between return and risk.

- The maximization problem can be written as:
$$
\arg \max_{\mathbf{w}} \frac{\mathbf{r}^T\mathbf{w} - r_f}
{\sqrt{\mathbf{w}^T\bm{\Sigma}\mathbf{w}}},~ \text{s.t.}~
\mathbf{w}^T\mathbf{1} = 1.
$$

- Rewrite it into a quadratic programming problem:
$$
\min_{\mathbf{w}} \mathbf{w}^T\bm{\Sigma}\mathbf{w},~
\text{s.t.}~ (\mathbf{r} - r_f \mathbf{1})^T \mathbf{w} = c_0,
$$
where $c_0$ is an arbitrary positive constant and $r_f$ is risk-free rate.

- Thus, the original
maximization problem is equivalent to the quadratic programming problem of
finding one weight vector $\mathbf{w}_k$, and scaling
$\mathbf{w}_k$ to $\mathbf{w}_0$ so that $\mathbf{w}^T_0\mathbf{1} = 1$.

## Efficient Frontier

- Following the framework originally developed by @markowitz1952, portfolios
maximizing expected return under certain given level of risk or equivalently
minimizing risk subject to a target expected return is defined as efficient
portfolios. The efficient frontier consists of all efficient portfolios.

    - The naive constructing method of efficient frontier is by obtaining the
      mean variance portfolios subject to different target returns.
      Alternatively, it can be shown that any minimum variance portfolio can
      be represented as a convex combination of any two minimum variance
      portfolios with different target expected returns.

    - If the expected return of the resulting portfolio is greater than the
      expected return of the global minimum variance portfolio, then the
      portfolio is an efficient frontier portfolio.

    - We can also get the efficient frontier based on the global minimum
      variance portfolio and tangency portfolio, both of which are known to
      be efficient.

## Simulation

- 20 assets.

- Each asset had 60 simulated rate of returns from normal distribution
with mean 0.005 and standard deviation 0.015.

- No specific correlation was imposed on these assets.

---

```{r efficient, echo = FALSE, fig.width = 6.5, fig.height = 3.5, fig.cap = fn("Sample efficient frontier."), fig.align = "center"}
## Tangency Portfolio ----------------------------------------------
## sample setting
na <- 20L                               # number of assets
ns <- 60L                               # number of observations
set.seed(1216)                          # important for reproducibility
retn <- matrix(rnorm(ns * na, mean = 0.005, sd = 0.015), nrow = ns)
mu <- colMeans(retn)                    # expected returns
rf <- 0.0001                            # riskfree rate (about 2.5% pa)
mu2 <- mu - rf                          # excess means
Q <- cov(retn)                          # covariance matrix

## obtain the optimized weights by quadratic programming
res_tan <- solve.QP(Dmat = Q, dvec = rep(0, na), Amat = matrix(mu2, ncol = 1),
                    bvec = 1, meq  = 1L)
w_tan <- prop.table(res_tan$solution)
retn_tan <- crossprod(w_tan, mu)
se_tan <- as.vector(sqrt(crossprod(w_tan, Q %*% w_tan))) # standard error
sr <- crossprod(w_tan, mu2) / se_tan  # compute sharpe ratio

## minimal variance portfolio
res_min <- solve.QP(Dmat = 2 * Q, dvec = rep(0, na),
                    Amat = matrix(1, nrow = na, ncol = 1),
                    bvec = 1, meq = 1L)
w_min <- res_min$solution
retn_min <- crossprod(w_min, mu)
se_min <- sqrt(as.vector(crossprod(w_min, Q %*% w_min)))

## mean-variance
eps <- 0.007
res_meanVar <- solve.QP(Dmat = 2 * Q, dvec = rep(0, na),
                        Amat = cbind(matrix(1, nrow = na, ncol = 1), mu),
                        bvec = c(1, eps), meq = 1L)
w_meanVar <- res_meanVar$solution
retn_meanVar <- crossprod(w_meanVar, mu)
se_meanVar <- sqrt(as.vector(crossprod(w_meanVar, Q %*% w_meanVar)))

## minimum variance with constrain
## eps <- 0.48
## w0 <- mu / sqrt(diag(Q))
## res_const <- solve.QP(Dmat = 2 * Q, dvec = rep(0, na),
##                       Amat = cbind(matrix(1, nrow = na, ncol = 1), w0),
##                       bvec = c(1, eps), meq = 1L)
## w_const <- res_const$solution
## retn_const <- crossprod(w_const, mu)
## se_const <- sqrt(as.vector(crossprod(w_const, Q %*% w_const)))

## construct efficient frontier from minimal variance and tangency portfolio
retnVec <- c(retn_min, retn_tan)
seVec <- c(se_min, se_tan)
varMat <- diag(seVec ^ 2)
varMat[1, 2] <- varMat[2, 1] <- as.vector(crossprod(w_min, Q %*% w_tan))
alpha <- seq(from = 1, to = - 1, by = - 0.1)
nAlpha <- length(alpha)
tmpMat <- cbind(alpha, 1 - alpha)
retn_eff <- tmpMat %*% retnVec
var_eff <- sapply(alpha, function(a) {
    crossprod(c(a, 1 - a), varMat %*% c(a, 1 - a))
})

## hue function
gg_color_hue <- function(n) {
    hues <- seq(15, 375, length = n + 1)
    grDevices::hcl(h = hues, l = 65, c = 100)[seq_len(n)]
}
colVec <- gg_color_hue(3)

## plot the tangency and minimal variace portfolio, on efficient portfolio
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.5, 0.5, 0)) # remove margin
plot(sqrt(var_eff), retn_eff, type = "l",
     xlab = expression(sigma[p]), ylab = expression(r[p]))
points(se_min, retn_min, col = colVec[1], pch = 20, cex = 1.5)
points(se_meanVar, retn_meanVar, col = colVec[2], pch = 20, cex = 1.5)
points(se_tan, retn_tan, col = colVec[3], pch = 20, cex = 1.5)
abline(b = (retn_tan - rf) / se_tan, a = rf, lty = 2, col = "gray", lwd = 1.5)
legend("bottomright", col = c("black", "gray", colVec),
       pch = c(NA, NA, rep(20, 3)), lty = c(1, 2, rep(NA, 3)),
       lwd = c(1, 1.5, rep(NA, 3)), cex = 0.85,
       legend = c("Efficient Frontier", "Captial Market Line (CML)",
                  "Global Minimum Variance", "Mean-Variance", "Tangency"))

```


# Backtesting

## Setup

- The sample stocks: *AAPL* (Apple Inc.), *AMZN* (Amazon Inc.),
  *GOOG* (Alphabet Inc.), and *FB* (Facebook Inc.).

- The weight of each stock was allocated at 10:30 AM on every market day.

- The trade was made weekly at 10:30 AM on the first day when the market
  opens in that week.

- For each asset, the trading weight was its average weight from last five
  daily allocations.

- The initial capital was set to be 10,000 USD. The time period for the
  backtesting was from 2014/06/01 to 2016/11/01.

- The only difference was the determination of weights of assets in the
  portfolio as discussed.


---


```{r fig:minVar, echo = FALSE, results = "asis", fig.align = "center"}
src_tag <- "figs/globalMinVar.png"
caption <- "Backtesting results of the global minimum variance portfolio."
lab_num <- figr("fig:minVar", TRUE, FALSE, type = "Figure")
caption_tag <- paste0(lab_num, ": ", caption)

if ("html" %in% outFormat) {
    cat(paste("<div class='figure' style='text-align: center'>",
              "<img src=", src_tag,
              "alt=", caption_tag, "/>",
              "<p class='caption'>", caption_tag, "</p></div>"))
} else {
    cat("![", caption, "][fig:globalMinVar]", sep = "")
}
```

---


```{r fig:meanVar, echo = FALSE, results = "asis", fig.align = "center"}
src_tag <- "figs/meanVariance.png"
caption <- "Backtesting results of the mean-variance portfolio."
lab_num <- figr("fig:meanVar", TRUE, FALSE, type = "Figure")
caption_tag <- paste0(lab_num, ": ", caption)

if ("html" %in% outFormat) {
    cat(paste("<div class='figure' style='text-align: center'>",
              "<img src=", src_tag,
              "alt=", caption_tag, "/>",
              "<p class='caption'>", caption_tag, "</p></div>"))
} else {
    cat("![", caption, "][fig:meanVar]", sep = "")
}
```

---


```{r fig:tangency, echo = FALSE, results = "asis"}
src_tag <- "figs/tangency.png"
caption <- "Backtesting results of tangency portfolio."
lab_num <- figr("fig:tangency", TRUE, FALSE, type = "Figure")
caption_tag <- paste0(lab_num, ": ", caption)

if ("html" %in% outFormat) {
    cat(paste("<div class='figure' style='text-align: center'>",
              "<img src=", src_tag,
              "alt=", caption_tag, "/>",
              "<p class='caption'>", caption_tag, "</p></div>"))
} else {
    cat("![", caption, "][fig:tangency]", sep = "")
}
```

# Summary

## Summary

- In summary, we reviewed the global minimum variance portfolio,
  mean-variance portfolio, and tangency portfolio learned from this course,
  and further compared their performance by back-testing in Quantopian.

- All of these portfolios outperformed the benchmark algorithm from
  Quantopian, though the performance of one certain portfolio is subject to
  the selection of stocks, follow-up time period, etc.

- The source document for this slide as well as the **Python** scripts used
  for backtesting in Quantopian are available at our GitHub repository
  named [math5800][github].


# Acknowledgment

## Acknowledgment

The completion of the project wouldn't be possible without the great effort
that our lovely team members have taken. It gave us much pleasure working
together as a group. We would also like to show our gratitude to Dr. Do who
introduced us the interesting lectures on financial programming and modeling,
and whose broad programming knowledge will have lasting effect.


# Reference

## Reference

[quantopian]: https://www.quantopian.com
[fig:meanVar]: figs/meanVariance.png
[fig:globalMinVar]: figs/globalMinVar.png
[fig:tangency]: figs/tangency.png
[github]: https://github.com/wenjie2wang/math5800
[Rmarkdown]: http://rmarkdown.rstudio.com/
